**Covid-19 vaccination in the province of Ontario: A geographical and socio-economical analysis**

### 2023-02-24 manuscript/01-background.qmd
working on discussion

The COVID-19 pandemic continues around the world with more than 600 million confirmed cases as of November of 2022 [@WHO-Covid]. During the first months of the pandemic in early 2020, non-pharmaceutical interventions such as masking and social distancing were the only methods available to manage the spread of the disease; however, the rapid development of vaccines permitted their approval and use in some countries towards the last month part of 2020. For example, in the US and Canada vaccine campaigns began in mid-December of 2020 [@tanne2020; @bogoch2022]. Although it has been estimated that vaccines against COVID-19 have prevented around 14 millions of deaths worldwide [@watson2022], the rollout of COVID-19 vaccines has faced multiple challenges since its inception.

>>>>

The COVID-19 pandemic continues around the world with more than 600 million confirmed cases as of November of 2022 [@WHO-Covid]. During the first months of the pandemic in early 2020, non-pharmaceutical interventions such as masking and social distancing were the only methods available to manage the spread of the disease; however, the rapid development of vaccines permitted their approval and use in some countries towards the last month part of 2020. For example, in the US and Canada vaccine campaigns began in mid-December of 2020 [@tanne2020; @bogoch2022]. Although it has been estimated that COVID-19 vaccines have prevented around 14 millions of deaths worldwide [@watson2022], the rollout of COVID-19 vaccines has faced multiple challenges since its inception.

### 2023-02-24 manuscript/03-results.qmd
working on discussion

@tbl-descriptive-stats shows the descriptive statistics (uncorrected) from the Fields Covid-19 survey data for vaccination status and each of the covariates analyzed. The total number of entries from the survey in the dataset after cleaning was 3,551. Overall, 26.9\% of survey respondents (958) reported not having received the first dose of the vaccine, whereas 73.1\% (2,593) reported having received it. Regarding socio-economic factors, respondents who reported living in a household with an income under CAD 15,000 or above 90,000 represented approximately 50\% of the total number of entries, with 71\% and 75\% of the respondents in each income bracket indicating to have received the first dose of the vaccine, respectively. The age group between 16-24 years had the highest representation in the survey responses (837, 23.6\% of all responses), followed by the age group of those between 65 and over (692, 19.5\% of the total); 73\% of those in the 16 to 24 years age group reported to have received the vaccine, whereas 68\% of those above 65 years indicated the same. The Health Region with the highest number of participants in the survey was Toronto, accounting for 1,342 entries (37.7\%) and a 72\% vaccination rate. Regarding race, individuals that identified as White/Caucasian represented 1313 (37\%) of all entries and had the highest vaccination uptake with 82\% of them indicating to have received the COVID-19 vaccine. On the other hand, those that identified as Indigenous had the lowest vaccine uptake with 60\% indicating to have received the vaccine.

@@

@tbl-model shows the results of the logistic regression on vaccination status using socio-economic, geographical factors (by Health Region) and their interactions as predictors. The reference groups in each case were set as follows: 16 to 34 years (age group), White Caucasian (Race), Toronto (Health Region), CAD 60,000 and over (Income). Statistically significant odds ratios were identified for individuals with a household income under CAD 25,000 (OR=0.37) and between CAD 25,000-59,999 (OR=0.59), persons who identified as Arab/Middle Eastern (OR=0.31), Black (OR=0.32), Latin American (OR=0.28), Other Race/Ethnicity (a group those that identified as Southeast Asian, Filipino, West Asian, and Minorities Not Identified Elsewhere, OR=0.22), and inviduals living in the West Health Region (OR=1.55).

Moreover, statistically-significant higher odd ratios were determined in the case of income and race; specifically, for individuals with a household income below CAD 25,000 that identified as Arab/Middle Eastern (OR=3.05), Black (3.19), Latin American (OR=2.80), and Other (4.59); and for individuals with a household income between CAD 25,000 and 59,999 that identified as Other racial minority (OR=6.93).

For Health Region and race, significant lower vaccination odds were identified for Black individuals in the Central Health Region (OR=0.44), and Other racial minorities and South Asian individuals that live in the West Health Region (ORs=0.41).

>>>>

@tbl-descriptive-stats shows the descriptive statistics (uncorrected) from the Fields Covid-19 survey data for vaccination status and each of the covariates analyzed. The total number of entries from the survey in the dataset after cleaning was 3,551. Overall, 26.9\% of survey respondents (958) reported not having received the first dose of the vaccine, whereas 73.1\% (2,593) reported having received it. Within each socio-economic factor, respondents who reported living in a household with an income under CAD 25,000 represented  37\% of the total number of entries, those within the CAD 25,000-59,999 income bracket represented 25% of the total sample, and those with an income above CAD 60,000 represented 38 \% of the sample; across all income brackets, the percentage of individuals that reported having received a first dose of the vaccine was consistent, above 69\%.

Within the age groups of survey respondents, the age group between 16-34 years had the highest representation in the survey responses (1,521, 42.8\% of all responses). Within this age bracket, 73\% of respondents indicated having received the vaccine, whereas the lowest vaccination rate was in the bracket of those 55 years of age and above, with a total of 72\%. The Health Region with highest representation in the survey was Toronto, accounting for 1,324 entries (37.2\%), with a vaccination rate of 72\%. Regarding race/ethnicity, individuals that identified as White/Caucasian represented 1313 (37\%) of all entries and had the highest vaccination uptake with 82\% of them indicating to have received the COVID-19 vaccine. On the other hand, the ethnic group with the lowest number of entries in the survey was Latin American, with a total of 180, or 5\% of all entries. Vaccination rates across all minority groups were below the value reported by White/Caucasians, with the lowest vaccination rate (60\%) being reported by individuals that identified as Indigenous.

@@

@tbl-model shows the results of the logistic regression on vaccination status using socio-economic factors (age group, income, race), geographical areas (Health Regions) and the interactions between income and race and Health Region and race. The reference groups were set as follows: 16 to 34 years (age group), White Caucasian (Race), Toronto (Health Region), CAD 60,000 and over (Income). There were no statistically significant differences in vaccination rates within the age groups, but significant odds ratios were estimated for individuals within the rest of the covariates. Within household income brackets, individuals with an income under CAD 25,000 or between CAD 25,000-59,999 had significantly lower odds of vaccination than those with an income above CAD 60,000 (ORs=0.37 and 0.59, respectively). Within Race/Ethnicity, individuals who identified as Arab/Middle Eastern, Black, or Latin American, had significantly lower odds of vaccination than those in the White/Caucasian group (ORs=0.31, 0.32, 0.28, respectively); additionally, those individuals in the Other Race/Ethnicity group (a group that included Southeast Asian, Filipino, West Asian, and Minorities Not Identified Elsewhere) had even lower odds of vaccination than the other minority groups, with an OR of 0.22. Regarding Health Regions, individuals that reported living in the West Health Region (which comprises the regions of Waterloo and Niagara, the counties of Wellington, Essex, and Lambton, and the cities of Hamilton, Haldimand, Brant, and Chatham-Kent) had significantly higher odds of vaccination than those in the Health Region of Toronto (OR=1.55).

Moreover, statistically-significant higher odd ratios were determined in the case of the interaction of income and race; specifically, for individuals with a household income below CAD 25,000 who identified as Arab/Middle Eastern (OR=3.05), Black (OR=3.19), Latin American (OR=2.80), or that belonged to other minority groups  (OR=4.59). Within the CAD 25,000-59,999 income bracket, individuals who identified as belonging to other racial minority groups had significantly higher odds of vaccination (OR=6.93).

For the interaction of Health Region and race, significant lower vaccination odds were identified for Black individuals in the Central Health Region, which comprises the region of York, counties of Dufferin and Simcoe and the district of Muskoka (OR=0.44). Finally, individuals that identified as part of other racial minorities or South Asian iand that lived in the West Health Region had significantly lower odds of vaccination  (ORs=0.41).

### 2023-02-24 manuscript/04-discussion.qmd
working on discussion

The rapid development of COVID-19 vaccines has been considered as a major achievement of modern medicine [@davis2022]. The availability of vaccines at the end of 2020 made some believe that they would be a determinant factor in a rapid ending of the pandemic [@thelancet2021]. However, despite previous successful vaccination campaigns that were crucial to control diseases such as smallpox and polio [@kayser2021], vaccination efforts in the case of Covid-19 have faced multiple challenges that have complicated the achievement of global immunity.

@@

In addition, vaccine uptake is influenced by geography as shown by different studies that have identified intra-regional differences in vaccine uptake [@pallathadka2022;@huang2022;@mollalo2021]. However, in the case of Canada, most studies that have analyzed spatial differences in vaccination have been focused in country-wide or province-wide estimates [@lavoie2022;@guay2022]. Therefore, we explored spatial and socio-economic determinants of vaccination status in the province of Ontario. This province is of particular interest as it has seen recently major structural health changes with the dissolution of the Local Health Integrated Networks (LHINs) and their incorporation into larger Health Regions [@dong2022]. The idea behind the change aimed at reducing the inequalities in healthcare that were identified under the LHIN approach [@tsasis2012], and thus, examining if there are significant differences in vaccination between the Health Regions could be important to consider by decision-makers in the development of public health policies that aim to address intra-provincial health disparities.

Our results indicate that across the most densely populated Health Regions of Ontario, almost three quarters of the surveyed individuals reported having received the first dose of the Covid-19 vaccine (@tbl-descriptive-stats). This is consistent with overall vaccination rates reported for Canada, which have been relatively higher when compared to other high income countries [@dube2022]. However, we also identified differences in vaccination likelihood based on different socio-economic factors. More specifically, our results show that those that were above 65 years old were only 44% likely to have received a vaccine when compared to those between 16 and 24 years old, that individuals with a household income below CAD 39,999 were around 60% less likely of having received the vaccine when compared to individuals in high income households (with an income above 90,000 CAD), and that individuals who identified as Arab and Middle Easterns, Black, Indigenous, Latin American, South Asian, Southeast Asian, Filipino, West Asian, and minorities not identified elsewhere had significantly lower probability of vaccination  (ranging between 15% and 40%) when compared to White/Caucasian individuals (@tbl-model).

These results are consistent with other studies that have shown an association between lower vaccination rates in racial minorities and lower household income [@guay2022;@muhajarine2021;@carter2022;@hussain2022]. However, it is interesting to note reduced vaccination likelihood in those aged 65 and over, as other studies have reported higher vaccination acceptance in this age group [@macdonald2021;@guay2022].

>>>>

The rapid development of COVID-19 vaccines has been considered as a major achievement of modern medicine [@davis2022], and their availability towards the end of 2020 in certain countries made some believe that they would be a determinant factor in a rapid ending of the pandemic [@thelancet2021]. However, despite previous successful vaccination campaigns that were crucial to control diseases such as smallpox and polio [@kayser2021], vaccination efforts in the case of Covid-19 have faced multiple challenges that have complicated the achievement of global immunity.

@@

In addition, vaccine uptake is influenced by geography as shown by different studies that have identified intra-regional differences in vaccine uptake [@pallathadka2022;@huang2022;@mollalo2021]. However, in the case of Canada, most studies that have analyzed spatial differences in vaccination have been focused in country-wide or province-wide estimates [@lavoie2022;@guay2022]. Therefore, we explored spatial and socio-economic determinants of vaccination status in the province of Ontario. This province is of particular interest as it has seen recently major structural health changes with the dissolution of the Local Health Integrated Networks (LHINs) and their incorporation into larger Health Regions [@dong2022]. The idea behind the change aimed at reducing the inequalities in healthcare that were identified under the LHIN approach [@tsasis2012], and thus, examining differences in vaccination between the Health Regions can provide decision makers with information regarding intra-provincial health disparities that may need to be addressed in the future.

Our results indicate that across the most densely populated Health Regions of Ontario, almost three quarters of the surveyed individuals reported having received the first dose of the Covid-19 vaccine (@tbl-descriptive-stats). This is consistent with overall vaccination rates reported for Canada, which have been relatively higher when compared to other high income countries [@dube2022]. However, we identified differences in vaccination based on socio-economic and geographical factors. First, our results show significant differences in vaccination odds in individuals with a household income below CAD 60,000, and in individuals belonging to minority racial or ethnic groups. Those who identified as Black, Latin American, or  belonging to different minorities not included as categories in the survey (Southeast Asian, Filipino, West Asian, and Minority not identified elsewhere) had vaccination odds below 33% when compared to individuals that identified as White/Caucasian (@tbl-model). Geographically, significantly higher vaccination odds were identified in the West Health Region, which comprises the regions of Waterloo and Niagara, the counties of Wellington, Essex, and Lambton, and the cities of Hamilton, Haldimand, Brant, and Chatham-Kent.

The lack of difference in vaccination uptake across different age groups is consistent with other studies that have shown similar results [@macdonald2021;@guay2022]. In the case of Ontario, where there was around 90% of vaccination uptake in the age group above 65 years old, and xxx percentages in xxx group and xxx group, these results show an overall high vaccination uptake at the provincial level.
These results are consistent with other studies that have shown an association between lower vaccination uptake, identifying as part of a racial minority, or having a low household income [@guay2022;@muhajarine2021;@carter2022;@hussain2022].
